Fast gravitational wave parameter estimation with no-compromise

### 2023-01-22 src/tex/ms.tex
edits through the end

% Brief description of PE
Parameter estimation (PE) is one of the most commonly performed data analysis
tasks in gravitational-wave (GW) data analysis and it underpins all of GW
physics and astrophysics \cite{Christensen:2022bxb, 2019PASA...36...10T}. The
central goal of PE is to infer the parameters of a particular GW source given
the strain data detected by instruments like LIGO \cite{LIGOScientific:2014pky},
Virgo \cite{VIRGO:2014yos} and KAGRA \cite{KAGRA:2020tym}. In the standard
compact binary coalescence (CBC) scenario, this could mean inferring intrinsic
parameters such as the masses and spins of the compact objects, as well as
extrinsic parameters such as their sky localization and distance from Earth. PE
is also applied to test general relativity (GR) and constrain deviations away
from its predictions in observed data
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary
evolution \cite{LIGOScientific:2021psn} and measuring the properties of nuclear
matter.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \cite{Veitch:2014wba}, \textsc{PyCBC Inference}
\cite{Biwer:2018osg}, and \textsc{Bilby} \cite{Ashton:2018jfp}. These packages
have been tested by a number of groups and are well regarded as the standard
tools. However, while these tools have passed many robustness tests, they are
known to be computationally intensive. The exact amount of time needed to
analyze one event depends on factors like the duration and frequency of the
signal, as well as features of the specific waveform model. Typical runtimes for
production-level analyses can range from days to weeks. This expense precludes
iterating quickly on results, launching large-scale measurement simulations, or
obtaining results in low latency to inform astronomers for potential follow-up in
real time.

@@

% Effort going to new tools development.
To address this, there are efforts from multiple groups developing
tools to speed up the PE process. This includes methods that employ modern
tools such as deep learning-based methods pre-trained on a large collection of
waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that simplify the
computational challenge by leveraging our knowledge of GW signals
\cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising avenues
to tackle standard GW analysis tasks in the coming decade, particularly about
events involving CBCs in GR, they rely on assumptions that may not be valid for
analysis involving additional physical effects such as lensing and beyond-GR
analysis, or may use approximations that do not compute the Bayesian likelihood
exactly.

@@

In this work, we present a lightweight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following major features to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific assumptions
about the problem to achieve its performance. This makes our method extensible
to problems beyond the standard CBC analysis, without sacrificing accuracy for
efficiency.

@@

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform model we use can be evaluated at any given time or frequency
independently, this means computing a waveform can be trivially parallelized
over frequency bins. Together with the heterodyned likelihood, we can evaluate
the likelihood at $\mathcal{O}(10^7)$ different locations on an Nvidia A100 GPU.
The high throughput of likelihood evaluations unlocks the potential of the
\texttt{flowMC} sampling algorithms.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
\cite{arviz_2019}} for GW150914 and GW170817 respectively. Running on an Nvidia A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in
%\kw{\variable{data/} and \variable{script}}
.

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW150914.png}
\caption{
    GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW150914}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{static/GW170817.png}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW170817}
\end{figure}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\cite{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
shown in \cite{release}.  We carry out these runs using \textsc{parallel Bilby}
(\textsc{pBilby}) \cite{Smith:2019ucc} to distribute the computation over 400
Intel Skylake CPUs for each event.  For the specific settings chosen, the
wall-time duration of each run is ${\sim}1$ day for GW150914 and ${\sim}2$ days
for GW170817.

@@

% Brief summary
In this work, we present a PE pipeline for GW events that is efficient and
flexible. Our package put together a number of innovations, including
differentiable waveform models from \texttt{ripple} \kw{cite}, heterodyned
likelihood and normalizing flow enhanced sampler \texttt{flowMC}. We tested the
robustness of our pipeline on a set of 1200 synthetic GW events, showing it is
capable of handling catalogs that will be available in the near future. We also
demonstrate our pipeline can estimate the parameters of GW150914 and GW170817 in
\kw{Quote speed}.

% Compared to other works
There are recent works from different groups on speeding up parameter
estimation of gravitational wave, including efficient reparameterization
\cite{Islam:2022afg,Roulet:2022kot} and deep learning
\cite{Dax:2021tsq,Dax:2022pxd}. While all of these methods can get down to
minutes-scale parameter estimation with high fidelity, our approach possesses
unique strengths, and may complement some of those other techniques.

% Stephen's work
Compared to \cite{Dax:2021tsq,Dax:2022pxd}, we do not rely on pre-training the neural
network on a large collection of waveforms and noise realization. This means as
soon as new waveform models and noise models are available, our algorithm can
already be deployed. Furthermore, our method is essentially an MCMC algorithm,
which inherits the merit of convergence measures in MCMC. As we are only using
the normalizing flow as a proposal distribution, and the normalizing flow is
trained jointly with a local sampler, we do not suffer from overfitting since
our training data is being generated on the fly and is always approaching the
target distribution. In this sense, we do not introduce potential extra
systematic error to inference results.

While our pipeline uses the sample generated by the local sampler for training,
one can also supply a pre-trained normalizing flow model to our pipeline to
bypass the training stage. This can further reduce the total runtime of our
inference pipeline. However, this may introduce potential systematic bias in the
inference result if the pre-trained network is not able to capture the
complexity presented in the data.

%Tousif's work
Compared to \cite{Islam:2022afg,Roulet:2022kot}, we do not rely on handcrafted
reparameterization of the coordinate systems. If the reparameterization scheme
is known ahead of time, a nice reparameterization is also encouraged. However,
handcrafted reparameterization depends on the assumptions used in deriving the
reparameterization scheme, which would inevitably run into limitations of use
cases. Our work can be viewed as an automatic reparameterization powered by
normalizing flow, while not being exact hence as efficient as an explicit
reparameterization, the method proposed in this work applies to a much border
class of problems, such as parameter estimation with precessing waveforms,
testing GR, and multi-event joint inference.

It is always beneficial to reparameterize if the reparameterization is known
ahead of time. For the class of problems where
\cite{Islam:2022afg,Roulet:2022kot} applies, we can incorporate the
reparameterization scheme into our MCMC pipeline to reduce the complexity of the
problem, hence speeding up the training phase. On the other hand, if there is
part of the target posterior that is not included in the reparameterization
scheme, the normalizing flow should still be able to learn to produce accurate
samples efficiently.

The two relevant works represent two orthogonal directions one can take in
building next generation tools in general. On one hand, there are modern tools
such as deep learning that are very flexible and powerful, but may need to rely
on having highly robust training data. On the other hand, there are traditional
tools that make use of our understanding of the underlying physics to simplify
the problem, which relies on having good intuition of the problem. Both
approaches rely on having a somewhat reasonable prior on how to approach the
problem. The main difference between methods used in industrial products and
scientific problems is scientific problems often try to address questions one
may have not been answered before, hence it is supposed to be far away from our
prior knowledge, meaning one need to take capability to generalize beyond the
current problem into account while building the tool. Our work utilizes both
reparameterization and deep learning, yet our method can be trivially extended
to problems beyond standard GW analysis that reparameterization or deep learning
alone may have trouble dealing with. We believe such flexibility is crucial for
building scientific tools.

% Additional traits for having differentiable waveforms
There are some future developments we are working on. While IMRPhenomD
\cite{Khan:2015jqa} is a reasonable start, it lacks some qualitative features
that other state-of-the-art models have, such as precession, higher mode, and
eccentricity. It also has a higher mismatch with reference numerical relativity
waveforms compared to more recent waveform models. Currently, we are working on
building differentiable IMRPhenomPv2 \cite{Khan:2018fmp} and three-parameters
NRSurrogate waveforms \cite{Varma:2018mmi}. Going forward, we encourage the
waveform development community to leverage autodiff environments such as Jax
when constructing new waveform models. Having a differentiable waveform model is
not only beneficial for parameter estimation, but also for other applications
such as calibrating waveforms to numerical relativity results, as detailed in
\kw{cite}.

% Access to higher dimensional problem
While standard GW analysis goes up to 17 dimensions. non-standard GW PE problems
can have more parameters, which could potentially lead to more complicated
geometry in the target posterior that is hard to reparameterize. For example,
\cite{LIGOScientific:2021sio} introduces 10 extra parameters to modify the
post-Newtonian coefficients predicted in GR, which should be taken care at the
same time during inference. On top of the increase of dimensionality, these
parameters often introduce non-trivial degeneracies such as local correlation
and multi-modality. Therefore, currently testing GR is limited to varying these
modifications one at a time, partially due to the bottleneck in the sampler.
Given the gradient-based and normalizing flow-enhanced sampler, our code shows
the potential in solving this problem in full at once.

% Other detectors configuration

To perform realistic PE with our pipeline, the antenna pattern of the
detector network needed to be taken into account. Our current code can perform
parameter estimation for any combination of the Hanford, Livingston, and Virgo
detectors, under the assumption of short signals such that the effect of Earth's
rotation can be ignored. For next-generation detector networks such as the
Einstein Telescope and the Cosmic Explorer, a differentiable version
of their antenna pattern is needed.

% More features such as marginalization
The features we have implemented are the barebone version of parameter
estimation. We do not include marginalization schemes such as time, phase, and
calibration lines marginalization \cite{2019PASA...36...10T}. Because of the reasonable performance of the
sampler on accelerators, time and phase marginalization is not necessary, as the
performance of our implementation is not significantly impacted by having two
extra dimensions. Nevertheless, having less parameters in a PE is always
appreciated, which we are looking into including in our code as well as other
marginalization modes.

% JIT overhead
Jax's JIT compilation drastically reduces the computational time to evaluate the
likelihood. However, it comes with a compilation overhead when the likelihood is
evaluated for the first time. We observe that the compilation time could depend
on the device where the code will be run. This is expected since Jax leverage
Accelerated Linear Algebra (XLA) takes advantage of accelerators, which means
Jax needs to compile the code for the specific device according to its
architecture. On an Nvidia A100 GPU, the compilation overhead could go up to 8
minutes for the waveform we are using. For the cases we have studied, the time
needed to obtain converging results on an A100 is about 2-3 minutes. This means
the compilation overhead is dominating the wall-clock time of the specific PE
run we considered. To utilize our implementation to its full potential, we are
looking into ways to reduce the compilation overhead or to cache the
compilation results to avoid paying the compilation overhead for every event.

% Faster maximum likelihood finder

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.

Another way to reduce the time of finding the maximum likelihood waveform is to
incorporate marginalization of extrinsic parameters to reduce the dimensionality
of the optimization problem. Currently, we simultaneously maximize all 11 GW
parameters, which is unnecessarily complicated and expensive. There are
long-existing marginalization schemes over extrinsic parameters such as the merger
time and phase, which can find the corresponding maximum likelihood waveform
much more efficiently when compared to differential evolution. We expect
implementing these marginalization schemes will reduce the time needed to find
the reference waveform parameters by obtaining the extrinsic parameters and by
reducing the dimensionality of the optimization problem for the intrinsic
parameters.

% Multi-device scaling
One important aspect of modern computing is scalability, meaning it is generally
favorable if one can simply put more computing units in the same problem and
reduce the Wall time. In our case, this means we would like to use more than one
GPU for the same PE process. More GPUs can help in the following ways: first,
more GPUs means we can run more independent chains at the same time, which can
generate more samples faster. But more importantly, as shown in this work and
\kw{cite flowMC?}, more independent chains also help with reducing the burn-in
time. Parallelizing over the number of chain dimension is trivial and does not
require much change to the current infrastructure. Another way more GPUs can
helps is by allowing faster training of larger flow models. While the training time
is not the biggest bottleneck given the flow model used in this study, more GPUs
means we can increase the bandwidth of the flow model by increasing its size
while keeping the training time the same. This helps capture more complex
geometry in the target space, which can lead to more accurate results in general.

\section{Acknowledgements}
Thanks Will.

>>>>

% Brief description of PE
Parameter estimation (PE) underpins all of gravitational-wave physics and
astrophysics, and is one of the most commonly performed tasks in
gravitational-wave (GW) data analysis \cite{Christensen:2022bxb,
2019PASA...36...10T}. The central goal of PE is to infer the parameters of a
particular GW source given the strain data recorded by instruments like LIGO
\cite{LIGOScientific:2014pky}, Virgo \cite{VIRGO:2014yos} and KAGRA
\cite{KAGRA:2020tym}.  In the standard compact binary coalescence (CBC)
scenario, this could mean inferring intrinsic parameters such the masses and
spins of the compact objects, as well as extrinsic parameters such as their sky
localization and distance from Earth. PE is also applied to test general
relativity (GR) and constrain deviations away from its predictions in observed
data
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}. PE
is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary
evolution \cite{LIGOScientific:2021psn} and measuring the properties of nuclear
matter.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \cite{Veitch:2014wba}, \textsc{PyCBC Inference}
\cite{Biwer:2018osg}, and \textsc{Bilby} \cite{Ashton:2018jfp}.  These packages
have been tested by a number of groups and are well regarded as standard tools.
However, while these tools have passed many robustness tests, they are known to
be computationally intensive. The exact amount of time needed to analyze one
event depends on factors like the duration and frequency of the signal, as well
as features of the specific waveform model. Typical runtimes for
production-level analyses can range from hours to weeks. This expense precludes
iterating quickly on results, launching large scale measurement simulations, or
obtaining results in low latency to inform astronomers for potential followup
in real time.

@@

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups to speed up
the PE process. This includes methods that employ modern tools such as methods
based on deep learning networks pre-trained on large collections of waveforms
\cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that reduce the
computational expense of classical PE by leveraging our knowledge of GW signals
\cite{Islam:2022afg,Roulet:2022kot}. While these methods are promising avenues
for standard GW problems, particularly for CBCs in GR, they rely on assumptions
that may not hold for analyses involving additional physical effects such as
lensing and deviations from GR, or may use approximations that do not compute
the Bayesian likelihood exactly.

@@

In this work, we present a lightweight, flexible, and high performance
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following techniques to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item differentiable waveform models,
\item normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler,
\item heterodyned likelihood,
\item native support for hardware accelerators.
\end{enumerate}
The main advantage of our framework is it does not rely on specific assumptions
about the problem to achieve its performance. This makes our method extensible
to problems beyond the standard CBC analysis, without sacrificing accuracy for
efficiency.

@@

To obtain a reference waveform, we currently use the \texttt{differential
evolution} algorithm \cite{Storn1997DifferentialE} available in the
\textsc{scipy} package \cite{2020SciPy-NMeth} to find the waveform parameters which maximize the
likelihood.
\mi{The reference waveform could also be produced from trigger parameters
precomputed by a search pipeline without additional computation.}

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform model we use can be evaluated at any given time or frequency
independently, this means computing a waveform can be trivially parallelized
over frequency bins. Together with the heterodyned likelihood, we can evaluate
the likelihood at $\mathcal{O}(10^7)$ different locations on an \textsc{Nvidia}
A100 GPU.  The high throughput of likelihood evaluations unlocks the potential
of the \textsc{flowMC} sampling algorithms.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
for the GW150914 analysis, and 128 s of data sampled at 4096 Hz for the
GW170817 analysis; strain data and PSDs for both events are fetched from GWOSC
\cite{GWOSC}. For our specific choice of sampler settings, we produce
${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using
\textsc{arviz} \cite{arviz_2019},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 10 minutes. Most of this
time is spent on just-in-time (JIT) compilation of the code; the actual
sampling time is only ${\sim}150$ s. We pre-compute the reference waveform
parameters used in hetetrodyne likelihood for the two events, the time for
which is omitted in the \texttt{wall} time calculation.  The chain data and the
analysis scripts which generate the chains can be found in
%\kw{\variable{data/} and \variable{script}}
.

\begin{figure}
    \script{plotEvents.py}
    \includegraphics[width=0.99\linewidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange).
    }
    \label{fig:GW150914}
\end{figure}

\begin{figure}
    \script{plotEvents.py}
\includegraphics[width=0.99\linewidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW170817}
\end{figure}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\cite{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
shown in \cite{release}.  We carry out these runs using \textsc{parallel Bilby}
(\textsc{pBilby}) \cite{Smith:2019ucc} to distribute the computation over 400
Intel Skylake CPUs for each event.  For the specific settings chosen, the
wall-time duration of each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day
for GW170817.

@@

\subsection{Comparison to other approaches}

% Compared to other works
There have been several recent efforts to speed up parameter estimation of
gravitational wave, relying on techniques ranging from efficient
reparameterizations \cite{Islam:2022afg,Roulet:2022kot} to deep learning
\cite{Dax:2021tsq,Dax:2022pxd}. While all of these methods can achieve
minutes-scale parameter estimation with high fidelity, our approach possesses
unique strengths, and may complement some of those other techniques.

% Stephen's work
In contrast to \cite{Dax:2021tsq,Dax:2022pxd}, we do not require pre-training
the neural network on a large collection of waveforms and noise realizations.
This means that our algorithm can be immediately deployed as soon as new
waveform models and noise models are available. Furthermore, our method is at
its core an MCMC algorithm, meaning it inherits the merit of convergence
measures in MCMC. As we are only using the normalizing flow as a proposal
distribution, and the normalizing flow is trained jointly with a local sampler,
we do not risk overfitting since our training data is being generated on the
fly and is always approaching the target distribution. In this sense, we do not
introduce potential extra systematic errors to the inference results.

While our pipeline uses samples generated by the local sampler for training,
one could also supply a pre-trained normalizing flow to our pipeline to bypass
the training stage. This would have the advantage of further reducing the total
runtime; however, it could introduce systematic bias in the inference result if
the pre-trained network is not able to capture the complexity presented in the
data.

%Tousif's work
In contrast to \cite{Islam:2022afg,Roulet:2022kot}, we do not rely on
handcrafted reparameterizations of the coordinate systems used for sampling. If
a useful reparameterization scheme is known ahead of time, it could trivially
be implemented within our pipeline, potentially easing convergence. However,
handcrafted reparameterizations rely on specific assumptions about the targeted
signal, which cannot always be generalized beyond specific applications. On the
other hand, within our pipeline, the normalizing flow effectively discovers
reparameterizations that ease sampling automatically without \emph{a priori}
knowledge of the structure of the problem. In general, the transformation
discovered by the normalizing flow will only be approximate and hence not as
efficient as an explicit reparameterization of the problem; yet, our approach
applies to a much border class of problems where clever reparameterizations are
not known ahead of time, such as parameter estimation with precessing
waveforms, calibration parameters, testing GR, and multi-event joint inference.

It is always beneficial to reparameterize if the reparameterization is known
ahead of time. For the class of problems treated in
\cite{Islam:2022afg,Roulet:2022kot}, we can incorporate those
reparameterizations directly into our MCMC pipeline to reduce the complexity of the
problem, hence speeding up the training phase. If there are limitations to the
reparameterization that mean it cannot properly encompass part of the target
posterior, the normalizing flow should still be able to learn to produce
accurate samples efficiently.

The two avenues discussed here (machine learning and reparameterizations)
represent two orthogonal directions one can take in building next generation PE
tools. On the one hand, there are modern techniques such as deep learning that
are very flexible and powerful, but rely on having highly robust training data.
On the other hand, there are traditional tools that make use of our
understanding of the underlying physics to simplify the problem, which relies
on having good intuition of the problem. Both approaches rely on having a
somewhat reasonable prior to approach the problem. The main difference between
methods used in industrial products and scientific problems is scientific
problems often try to address questions one may have not been answered before,
hence it is supposed to eventually depart from our prior knowledge, meaning its
desirable to design methods that generalize beyond the current state of
knowledge robustly. Our work utilizes both reparameterization and machine
learning, yet our method can be trivially extended to problems beyond standard
GW analysis that reparameterization or deep learning alone may have trouble
dealing with. Beyond efficiency, we believe such flexibility and robustness are
crucial for building scientific tools.

\subsection{Future development}

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other
state-of-the-art models have, such as precession, subdominants moments of the
radiation, and eccentricity. It also has a higher mismatch with reference
numerical relativity waveforms compared to more recent waveform models.
Currently, we are working on building differentiable implementations of
\textsc{IMRPhenomPv2} \cite{Khan:2018fmp}, the precessing successor to
\textsc{IMRPhenomD}, as well as the numerical relativity surrogate waveforms,
including \textsc{NRSur7dq4} \cite{Varma:2018mmi}. Going forward, we expect the
use of autodifferentiation environments like \textsc{Jax} \cite{} to become
more prevalent in the waveform development community, increasing the number of
differentiable waveform models available. This would not only be beneficial for
parameter estimation, but also for a number of other applications such as
Fisher matrix computations, template placement and calibrating waveforms to
numerical relativity \kw{cite}.

% Access to higher dimensional problem
While standard CBC analyses go up to 17 dimensions, non-standard GW PE problems
can have more parameters, which could potentially lead to more complicated
geometry in the target posterior that is hard to reparameterize. For example,
\cite{LIGOScientific:2021sio} introduces 10 extra parameters controlling
deviations in the post-Newtonian coefficients predicted in GR. On top of the
increase of dimensionality, these parameters often introduce non-trivial
degeneracies such as local correlation and multi-modality. Therefore, currently
testing GR is limited in practice to varying these modifications one at a time,
partially due to the bottleneck in the sampler.  Given the gradient-based and
normalizing flow-enhanced sampler, our code shows promise in tackling this
problem.

% Other detectors configuration
Our current code can perform parameter estimation for any combination of
ground-based detectors detectors, under the assumption that signals are
transient and their wavelength is short. The first condition guarantees that
the effect of Earth's rotation can be ignored when computing antenna patterns,
while the second means that we can treat the antenna patterns as frequency
independent constants. These assumptions break for next-generation detectors,
whether on Earth or in space, like Cosmic Explorer, Einstein Telescope and
LISA; differentiable implementations antenna patterns for those detectors is
work in progress.

% More features such as marginalization
Furthermore, our current implementation is minimal and we do not make use of
most standard ``tricks'' to accelerate sampling.  In particular, we do not
incorporate (semi)analytic marginalization schemes over parameters such as
time, phase, and calibration parameters \cite{2019PASA...36...10T}. Because of
the already reasonable performance of our sampler thanks to hardware
accelerators, time and phase marginalization is not crucial, as the performance
of our implementation is not significantly impacted by having two extra
dimensions. Nevertheless, having fewer parameters cannot hurt in the future, so
we are looking into incorporating analytic marginalization schemes in our code
as well.

% JIT overhead
When it comes to wall time, the just-in-time compilation of our code is the
current limiting factor.  While \textsc{Jax}'s JIT compilation drastically
quickens likelihood evaluations, it comes with a significant compilation
overhead before the first evaluation. We observe that the compilation time
depends on the device where the code is run; this is expected since
\textsc{Jax} leverages the \textsc{Accelerated Linear Algebra} (XLA) compiler
to take advantage of hardware accelerators, which means that \textsc{Jax} needs
to compile the code for each specific device according to its architecture. On
an \textsc{Nvidia} A100 GPU, the compilation overhead could go up to 8 minutes
for our current waveform. Meanwhile, for the cases we have studied, the time
needed to obtain converging sampling on an A100 is about 2-3 minutes. This
means the compilation overhead dominates the wall-clock time of our current PE
runs. To maximize the potential of our code, we are looking into ways to reduce
the compilation overhead or to cache the compilation results to avoid paying
the compilation overhead for every event.

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\texttt{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 5 minutes \kw{Benchmark it
in the code and make this number precise.} for GW170817. There are two ways to
reduce this time.

First, we can explore a different optimization strategy that
takes full advantage of the strengths of our pipeline, in particular the
differentiability of our likelihood and the possibility to evaluate many
waveforms in parallel with a GPU. Particle swarm \cite{7869491} and stochastic
gradient descent methods \cite{10.5555/304710.304720} are promising candidates
we are investigating.

Second, we may marginalize extrinsic parameters to reduce the dimensionality of
the optimization problem. Currently, we simultaneously maximize all 11 CBC
parameters in our problem numerically, which is unnecessary. There are
long-existing, efficient maximization schemes for extrinsic parameters, such as
the merger time and phase, which can find the corresponding maximum likelihood
waveform much more efficiently when compared to differential evolution. We
expect implementing these schemes will reduce the time needed
to find the reference waveform parameters by fixing the extrinsic parameters
and by reducing the dimensionality of the optimization problem.

% Multi-device scaling
Finally, one important aspect of modern computing is scalability, meaning it is
generally favorable if one can simply put more computing units in the same
problem and reduce the wall time. In our case, this means that we would like to
use more than one GPU for the same PE process. More GPUs can help in the
following ways: first, more GPUs means we can run more independent chains at
the same time, which can generate more samples faster; second, and more
importantly, as shown in this work and \kw{cite flowMC?}, more independent
chains also help with reducing the burn-in time. Parallelizing over the number
of chain dimension is trivial and does not require much change to the current
infrastructure. Additional GPUs can also help by enabling faster training of
larger flow models. While the training time is not the biggest bottleneck given
the flow model used in this study, more GPUs means we can increase the
bandwidth of the flow model by increasing its size while keeping the training
time the same. This would help capture more complex geometries in the target
space, which can lead to better convergence in general.

\section{Conclusion}

% Brief summary
In this work, we presented a PE pipeline for GW events that is efficient,
flexible and reliable. Our package brings together a number of innovations,
including differentiable waveform models, likelihood heterodyning, and
normalizing-flow enhanced gradient-based sampling. We tested the robustness of
our pipeline, currently built upon \textsc{ripple} and \textsc{flowMC}, on a
set of 1200 synthetic GW events, showing it is robust, unbiased and efficient
enough to handle the large catalogs of detections that will be available in the
near future. We also show that our pipeline can estimate the parameters of
GW150914 and GW170817 in \kw{Quote speed}, demonstrating the potential of our
implementation on real data.

\section{Acknowledgements}
We thank Will M.~Farr for illuminating discussions.
The Flatiron Institute is a division of the Simons Foundation, supported through the generosity of Marilyn and Jim Simons.
This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gw-openscience.org), a service of LIGO Laboratory, the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain. The construction and operation of KAGRA are funded by Ministry of Education, Culture, Sports, Science and Technology (MEXT), and Japan Society for the Promotion of Science (JSPS), National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea, Academia Sinica (AS) and the Ministry of Science and Technology (MoST) in Taiwan.
